Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2019-10-23 paper.tex
once again through intro + alg sections

@@

A \emph{ranked phylogenetic tree} is a pair consisting of a rooted binary phylogenetic tree $T$ on the set $X = \{1, \ldots, n\}$ of taxa for $n \in \mathbb N$, and a (total) rank function $\rank$ that maps all elements of $X$ to $0$, all internal nodes of $T$ onto the set $\{1, \ldots, n-1\}$, and respects the partial order given by the tree.
The latter means that if $x$ and $y$ are two internal nodes of $T$ such that there exists a path from a taxon $z \in X$ to $y$ which passes through $x$ then $\rank(x) < \rank(y)$.
Ranked trees $(T_1, \rank_1)$ and $(T_2, \rank_2)$ are different if trees $T_1$ and $T_2$ are different or $T_1 = T_2$ and $\rank_1 \neq \rank_2$.
Since all trees in this paper are ranked trees, we will abuse the notation and drop the ranking from the notation.
We will also refer to a ranked tree simply by tree.
For a tree $T$, we will use $\rank_T$ to refer to its ranking.

Our definition of a ranked tree implies a natural notion of the edge length.
Indeed, a tree edge $(u,v)$ has length $l$ if $|\rank(v) - \rank(u)| = l$.

@@

The vertex set of the $\rnni$ graph is the set of all ranked trees on $n$ taxa.
We introduce two types of operation ($\rnni$ \emph{moves}) on trees (see Figure~\ref{fig:RNNI}) and say that two trees are adjacent in the $\rnni$ graph is they are connected by an operation of either type.
The first type of operation is called a \emph{rank move} and defined by swapping the ranks of two internal nodes that are not adjacent in the tree.
Formally, if $x$ and $y$ are such nodes of a tree $T$ that $|\rank_T(x) - \rank_T(y)| = 1$ and $(x, y)$ is not an edge in $T$ then the tree $R$ obtained from $T$ by only changing $\rank_R(x) = \rank_T(y)$ and $\rank_R(y) = \rank_T(x)$ is said to be obtained by a rank move.
The second type of operation is called an $\nni$ \emph{move} and defined in the usual way, that is, two trees $T$ and $R$ are said to be connected by an $\nni$ move if there exist internal edges $e$ in $T$ and $f$ in $R$ both of length one such that the trees obtained by shrinking $e$ and $f$ to internal nodes coincide.

@@

\todo{Add a bit more space in all figures at the bottom to have at least a standard interval above he caption.}
\todo{In this figure -- rm arrows (think how to place the RNNI edges), possibly avoid using colour, and highlight where the rank move is made (hollow nodes?)}
\todo{Two of these trees are caterpillar trees.
Do you think it would be helpful to draw them as caterpillars to prepare readers for the rest of the paper?}
\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{RNNI}
	\caption{Two types of operation that define edges of the $\rnni$ graph.}
	\label{fig:RNNI}
\end{figure}

\todo{label of taxa in figure shouldn't be numbers\\
AG: Well -- if you decide to change this, go through all the definitions VERY carefully to make sure we update everywhere.}

@@

\todo{LC: make sure that we move all definitions in this paper to the right place}

Within this section we will first two introduce algorithms, namely $\findpath$ and $\csort$, for computing paths between two given trees in $\rnni$.
We will use them later to prove some geometric properties of the $\rnni$ graph in section~\ref{section:geometry}.
Additionally we will present the algorithm $\mdtree$, which computes a caterpillar tree with maximum distance
\todo{LC: can we use the term max dist here?}
from a given tree.

% In \autocite{Gavryushkin2018-ol} an efficient algorithm for computing the $\rnni$ graph has been introduced.
% We implemented this algorithm, which produces the graph for trees with up to $8$ taxa in reasonable time. \todo{Shall we say more about the algorithm here? + running time (in hours/minutes) on which computer?}
% This enables us to compute exact distances between trees with up to $8$ taxa.
% In particular, comparing exact distances with approximations as introduced in previous sections and verifying the Cluster Theorem of Section~\ref{section:split_theorem} can now be for small trees.
% All algorithms have been implemented with Python and are available online. \todo{link?}
% In the following, further details about the implementation of algorithms $\findpath$ and $\mdtree$ as well as a description of how to prove the Cluster Theorem will be given.

@@

\todo{LC: Introducing words}

\todo{LC: move the following definition? - it's actually quite important not only for the algorithm, so we might want to have it in the introduction?}
Before introducing the algorithm, we need the following definitions.
Each node $v$ of a tree defines a \emph{cluster} $C$, which is the set of taxa descending from $v$.
We then say that $v$ \emph{induces} $C$.
\emph{Trivial clusters} are clusters induced by leaves (contain just one element each) and the root of the tree (contains all taxa).
Note that the list of all
\todo{Check that the indexing of $C_i$ is still correct. ($i=1,\ldots,n-2$)}
non-trivial clusters of a tree, sorted according to the rank of the corresponding node, defines the tree unambiguously.
For the rest of this paper we will use the notion cluster and mean non-trivial clusters only because all trees on $n$ taxa share the trivial clusters.
\todo{LC: so we ever use trivial clusters? If so, we need to add a comment here}

We are now presenting an algorithm for computing paths in $\rnni$ between any two given trees.
Let $C_1, \ldots, C_{n-2}$ be the clusters of $R$ (the destination tree) where $C_i$ is induced by the node with rank $i$ for $i = 1, \ldots, n-2$.
Within every step of the algorithm we add one more tree to extend a path $p$, which will be the output path of $\findpath$.
First, set $\hat{T} = T$ to be the first element in $p$.
With $\hat T$ we denote the last tree of $p$ at each step in the algorithm.
Consider the clusters $C_1, \ldots, C_{n-2}$ of $R$ in this order.
When arriving at cluster $C_k$, find the node inducing the smallest cluster containing all elements of $C_k$ in $\hat T$.
This node is denoted by $\mrca_{\hat{T}}(C_k)$ (short for \emph{most recent common ancestor}).
\todo{LC: make sure you don't use mrca before this!}
The rank of $\mrca_{\hat{T}}(C_k)$ in $\hat T$ is higher than the rank of the node inducing $C_k$ in $R$, because we consider $C_i$ for increasing $i$.
As a result, all clusters $C_1, \ldots, C_{i-1}$ are induced by some nodes of $\hat T$ as they are in the destination tree $R$.
Now update $\hat{T}$ by performing an $\rnni$ move that decreases the rank of $\mrca_{\hat{T}}(C_i)$.
With Lemma~\ref{lemma:mrca_move} we know that there is always exactly one $\rnni$ move possible that decreases the rank of $\mrca_{\hat{T}}(C_i)$.
Repeat this until $\mrca_{\hat{T}}(C_i)$ has the same rank $i$ in $\hat T$ as in $R$.
Thus, repeating this procedure for all clusters of $R$ results in a path between any pair of tress $T, R$, indeed.

@@

With $\findpath$ we can compute a path between two arbitrary trees in $\rnni$.
The length of such a path can then be used as approximation for the distance between the trees.
We implemented $\findpath$ in Python.
\todo{LC: cite, link to github code?}
\todo{LC: Shall we mention the exact algorithm at the beginning of this section already?}
In~\autocite{Gavryushkin2018-ol} an algorithm for computing the $\rnni$ graph is being introduced.
We implemented this algorithm
\todo{LC: github?}
and used
\todo{LC: which algorithms did we use?}
to compute exact distances between all pairs of trees on up to $7$ taxa.
\todo{LC: Is 7 the actual number?}
By comparing these distances with the distances given by $\findpath$, we find that $\findpath$ gives the correct distances for all pairs of trees on up to $7$ taxa.
\todo{LS: Is it 7? How did we compute all distances with $\findpath$? - We used the symmetry of $\rnni$}
The running time of $\findpath$, implemented as described by the pseudo-code in Algorithm~\ref{alg:find_path}, is quadratic in the number of taxa.

\begin{algorithm}[H]
\caption{$\findpath$($T,R$)}
\label{alg:find_path}
\begin{algorithmic}[1]
	\STATE $\hat{T} := T$
	\STATE $p := (T)$
	\FOR {$i = 1, \dots, n-2$}
		\STATE Let $C$ be the cluster induced by node $v$ in $R$ with $\rank(v) = i$
		\WHILE {$\rank(\mrca_{\hat{T}}(C))>i$}
			\STATE Update $\hat{T}$: Decrease $\rank(\mrca_{\hat{T}}(C))$ by an $\rnni$ move \label{alg:line:move_set_down}
			\STATE $p = p+\hat{T}$
		\ENDWHILE
	\ENDFOR
	\RETURN $p$
\end{algorithmic}
\end{algorithm}

In Section~\ref{section:diameter} we will see that for any pair of trees, $\findpath$ computes a path that is at most as long as the the diameter of $\rnni$.
Thus, if the distance between two trees equals the diameter, $\findpath$ computes the correct distance.
\todo{LC: Can we say something about the approximation ratio here?}

\subsection{$\csort$}
\label{section:alg_cpath}

@@

Before we start introducing an algorithm for computing paths between caterpillar trees, we need to introduce further notations.
A path $p$ between two caterpillar trees $T$ and $R$ that only consists of caterpillar trees is called \emph{caterpillar path}.
If $p$ is shortest among all caterpillar paths between these trees we call it \emph{shortest caterpillar path}, and the length of $p$ will be denoted by $d_c(T,R)$ and called \emph{caterpillar distance}.

\todo{LC: structure in the following paragraph!}
We next introduce a unique way to represent of caterpillar trees as sequences of taxa.
We will use this to show how one can use the classical \emph{Bubble Sort} algorithm \autocite{Knuth1997-pi} to compute caterpillar paths.
This algorithm is denoted by $\csort$ and the pseudo-code is provided in Algorithm.
\todo{LC: pseudo-code}
One can see caterpillar trees as sequences of taxa, ordered non-decreasing according to the ranks of their parents.
Note that the only two elements in such a sequence that have parents of equal rank are the first two elements, the cherry taxa $c_1$ and $c_2$.
To make the sequence representation of caterpillar trees unique, we assume that for it is $c_1 < c_2$.
$\nni$ moves between caterpillar trees, which are the only possible $\rnni$ moves between those trees, coincide with exchanges of two neighboured elements in the sequence.
The only exception are the two first three elements in the sequence.
Corresponding to $\nni$ moves on the lowest internal edge are exchanges of either $c_1$ or $c_2$ with the third elements $c_3$ in the sequence.
But a swap of $c_1$ and $c_2$ does not correspond to any move in $\rnni$.
Therefore, we need to adapt Bubble Sort for the first three elements of the sequence as follows:
\todo{LC: add the pseudocode and refer to the line where $\csort$ differs from Bubble Sort! (We assume that we are at the point where we do some changes only on the first positions of the sequence)}
If the rank of $\parent_R(c_3)$ is higher than the ranks of both $\parent_R(c_1)$ and $\parent_R(c_2)$ in $R$, there are no swaps needed for these taxa.
If just one of the two taxa $c_1$ or $c_2$ is on the other side of $c_3$ in $R$, this taxon swaps position with $c_3$ in the sequence.
And if both $c_1$ and $c_2$ are on the other side of $c_3$ in $R$, the taxon that has the parent with higher rank in $R$ swaps with $c_3$.
This makes sure that there is no additional swap of $c_1$ and $c_2$ needed on the subsequent path to $R$.

@@

In this section we analyse shortest paths in $\rnni$.
We will start by presenting an efficient algorithm for computing (not necessarily shortest) paths in $\rnni$ ($\findpath$) that turns out to be an essential tool for understanding the geometry of $\rnni$.
Specifically, this algorithm will enable us to prove that there is always a shortest path between two caterpillar trees that only consists of caterpillar trees (Theorem~\ref{thm:caterpillar_convex}) in Section~\ref{section:caterpillar_convex}.
In other words, the set of caterpillar trees is \emph{convex} in $\rnni$.
\todo{The rest of this paragraph should be carefully checked and be in the introduction.
Generally, you might want to think a bit more about a better structure of this paper into chapters.}
This distinguishes $\rnni$ from $\nni$ where this does not hold.
One further result, presented in Section~\ref{section:diameter}, directly following from the proof of Theorem~\ref{thm:caterpillar_convex}, is the diameter of $\rnni$.
Finally, we disprove the so-called Strong Split Theorem, conjectured in \autocite{Gavryushkin2018-ol}.
However, the Weak Split Theorem remains a conjecture.
Furthermore, we propose an alternative (even weaker) version of the conjecture, the Cluster Theorem (Section~\ref{section:split_theorem}).

>>>>

\todo{LC: mention some results here and how they make $\rnni$ differ from $\nni$}
This distinguishes $\rnni$ from $\nni$ where this does not hold.
One further result, presented in Section~\ref{section:diameter}, directly following from the proof of Theorem~\ref{thm:caterpillar_convex}, is the diameter of $\rnni$.
Finally, we disprove the so-called Strong Split Theorem, conjectured in \autocite{Gavryushkin2018-ol}.
However, the Weak Split Theorem remains a conjecture.
Furthermore, we propose an alternative (even weaker) version of the conjecture, the Cluster Theorem (Section~\ref{section:split_theorem}).

@@

A \emph{ranked phylogenetic tree} is a pair consisting of a rooted binary phylogenetic tree $T$ on the set $X = \{1, \ldots, n\}$ of \emph{taxa} for $n \in \mathbb N$, and a (total) rank function $\rank$ that maps all elements of $X$ to $0$, all internal nodes of $T$ onto the set $\{1, \ldots, n-1\}$, and respects the partial order given by the tree.
The latter means that if $u$ and $v$ are two internal nodes of $T$ such that there exists a path from a taxon $x \in X$ to the root which first passes through $u$ and then through $v$ then $\rank(u) < \rank(v)$.
Ranked trees $(T_1, \rank_1)$ and $(T_2, \rank_2)$ are different if trees $T_1$ and $T_2$ are different or $T_1 = T_2$ and $\rank_1 \neq \rank_2$.
Since all trees in this paper are ranked trees, we will abuse the notation and drop the ranking from the notation.
We will also refer to a ranked tree simply by tree.
For a tree $T$, we will use $\rank_T$ to refer to its ranking.

Our definition of a ranked tree implies a natural notion of the edge length.
Indeed, a tree edge $(u,v)$ of $T$ has length $l$ if $|\rank_T(v) - \rank_T(u)| = l$.

@@

The vertex set of the $\rnni$ graph is the set of all ranked trees on $n$ taxa.
We introduce two types of operation ($\rnni$ \emph{moves}) on trees (see Figure~\ref{fig:RNNI}) and say that two trees are adjacent in the $\rnni$ graph if they are connected by an operation of either type.
The first type of operation is called a \emph{rank move} and defined by swapping the ranks of two internal nodes that are not adjacent in the tree.
Formally, if $u$ and $v$ are such nodes of a tree $T$ that $|\rank_T(u) - \rank_T(v)| = 1$ and $(u, v)$ is not an edge in $T$ then the tree $R$ obtained from $T$ by only changing $\rank_R(u) = \rank_T(v)$ and $\rank_R(v) = \rank_T(u)$ is said to be obtained by a rank move.
The second type of operation is called an $\nni$ \emph{move} and defined in the usual way, that is, two trees $T$ and $R$ are said to be connected by an $\nni$ move if there exist internal edges $e$ in $T$ and $f$ in $R$ both of length one such that the trees obtained by shrinking $e$ and $f$ to internal nodes coincide.

@@

\todo{Add a bit more space in all figures at the bottom to have at least a standard interval above he caption.}
\todo{In this figure -- rm arrows (think how to place the RNNI edges), possibly avoid using colour, and highlight where the rank move is made (hollow nodes?)}
\todo{Two of these trees are caterpillar trees.
Do you think it would be helpful to draw them as caterpillars to prepare readers for the rest of the paper?}
\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{RNNI}
    \vspace{2pt}
	\caption{Two types of operation that define edges of the $\rnni$ graph.}
	\label{fig:RNNI}
\end{figure}

@@

Within this section we will first introduce two algorithms, namely $\findpath$ (Section~\ref{section:alg_findpath}) and $\csort$ (Section~\ref{section:alg_csort}), for computing paths between two given trees in $\rnni$.
The first one takes arbitrary trees as input while the latter one is restricted to trees with a certain tree shape.
We will use them later to prove some geometric properties of the $\rnni$ graph in section~\ref{section:geometry}.
Additionally we will present the algorithm $\mdtree$, which produces a tree that is far away from the given input tree.
What we mean by far away depends on the input tree and will be discussed in Section~\ref{section:alg_mdtree}.

@@

Before introducing the algorithm we need the following definitions.
Each node $v$ of a tree defines a \emph{cluster} $C$, which is the set of taxa descending from $v$.
We then say that $v$ \emph{induces} $C$.
\emph{Trivial clusters} are clusters induced by leaves (contain just one element each) and the root of the tree (contains all taxa).
Note that the list of all non-trivial clusters of a tree, sorted according to the rank of the corresponding node, defines the tree unambiguously.
For the rest of this paper we will use the notion cluster and mean non-trivial clusters only because all trees on $n$ taxa share the trivial clusters.

We are now presenting $\findpath$ (Algorithm~\ref{alg:find_path}), an algorithm for computing paths between any two given trees in $\rnni$.
Let $C_1, \ldots, C_{n-2}$ be the clusters of $R$ (the destination tree) where $C_i$ is induced by the node with rank $i$ for $i = 1, \ldots, n-2$.
$\findpath$ computes a path $p$ which initially consists of $T$ only and is extended stepwise.
Let $\hat T$ denote the last tree of $p$ at any point in the algorithm.
For extending $p$, which equals updating $\hat T$, the clusters $C_1, \ldots, C_{n-2}$ are considered in this order.
When arriving at cluster $C_i$ the clusters in $\hat T$ are rearranged by some $\rnni$ moves until $C_i$ is induced by a node in $\hat T$.
This works as follows.
First find the node inducing the smallest cluster containing all elements of $C_k$ in $\hat T$.
This node is denoted by $\mrca_{\hat{T}}(C_k)$ (short for \emph{most recent common ancestor}).
The rank of $\mrca_{\hat{T}}(C_k)$ in $\hat T$ is higher than the rank of the node inducing $C_k$ in $R$, because we consider $C_i$ for increasing $i$.
As a result, all clusters $C_1, \ldots, C_{i-1}$ are induced by some nodes of $\hat T$ as they are in the destination tree $R$.
Now update $\hat{T}$ by performing an $\rnni$ move that decreases the rank of $\mrca_{\hat{T}}(C_i)$.
With Lemma~\ref{lemma:mrca_move} we know that there exists always exactly one $\rnni$ move that decreases the rank of $\mrca_{\hat{T}}(C_i)$.
So we can perform these moves until $\mrca_{\hat{T}}(C_i)$ has the same rank $i$ in $\hat T$ as in $R$.
Repeating this procedure for all clusters of $R$ results in a path from $T$ to $R$.

\begin{algorithm}[H]
\caption{$\findpath$($T,R$)}
\label{alg:find_path}
\begin{algorithmic}[1]
	\STATE $\hat{T} := T$
	\STATE $p := [\hat T]$
	\FOR {$i = 1, \dots, n-2$}
		\STATE Let $C_i$ be the cluster induced by the node with rank $i$ in $R$
		\WHILE {$\rank(\mrca_{\hat{T}}(C_i))>i$}
			\STATE Update $\hat{T}$: Decrease $\rank(\mrca_{\hat{T}}(C_i))$ by an $\rnni$ move \label{alg:line:move_set_down}
			\STATE $p = p+\hat{T}$
		\ENDWHILE
	\ENDFOR
	\RETURN $p$
\end{algorithmic}
\end{algorithm}

@@

Since $\findpath$ computes a path between two arbitrary trees in $\rnni$, we can use the length of such a path as approximation for the distance between the trees.
A natural question is now how well this estimates the actual distance in $\rnni$.
To answer this question for small trees, we implemented $\findpath$.
\todo{LC: github - link}
For computing the actual $\rnni$ distances we used the the algorithm for computing the $\rnni$ graph that was introduced in Section 3.3 of \autocite{Gavryushkin2018-ol}.
With our implementation of this algorithm
\todo{LC: gihub - link}
we can compute the graph for trees with up to $8$ taxa in reasonable time. \todo{LC: Shall we say more about the algorithm here? + running time (in hours/minutes) on which computer?}
We used
\todo{LC: How do we compare the distances?}

The running time of $\findpath$, implemented as described by the pseudo-code in Algorithm~\ref{alg:find_path}, is quadratic in the number of taxa.
\todo{LC: Where do we mention the running time?}

\subsection{$\csort$}
\label{section:alg_csort}

@@

Before introducing an algorithm for computing paths between caterpillar trees, we need to introduce further notations.
We aim to find paths between \emph{caterpillar trees} where each internal node has at least one child that is a leaf.
A path $p$ between two caterpillar trees $T$ and $R$ that only consists of caterpillar trees is called \emph{caterpillar path}.
If $p$ is shortest among all caterpillar paths between these trees we call it \emph{shortest caterpillar path}, and the length of $p$ will be denoted by $d_c(T,R)$ and called \emph{caterpillar distance}.

We next introduce a representation of caterpillar trees as sequences of taxa.
We will use it to show how one can adapt the classical \emph{Bubble Sort} algorithm \autocite{Knuth1997-pi} to compute caterpillar paths.
\todo{LC: pseudo-code}
Ordering the taxa of a caterpillar trees as a non-decreasing, according to the ranks of their parents, gives us a sequence of length $n$.
Note that the only two elements in such a sequence that have parents of equal rank are the first two elements $c_1$ and $c_2$.
We will call a pair of two taxa that share their parent \emph{cherry}.
\todo{LC: pair of taxa or subtree (includes internal node and two edges)}
To make the sequence representation of caterpillar trees unique, we assume that it is $c_1 < c_2$ for the first two elements in the sequence.

In the following we explain how $\rnni$ moves between caterpillar trees work in the sequence representation.
At first it is important to realise that the only possible $\rnni$ moves between caterpillar trees are $\nni$ moves, since there is no pair of internal nodes with rank difference one that is not connected by an edge.
$\nni$ moves between caterpillar trees correspond to swaps of two neighboured elements in the sequence representation.
The only exception of this are $\nni$ moves on the lowest internal edge, that is the edge connecting the internal nodes of rank one and two.
There are two $\nni$ moves possible on this edge that result in caterpillar trees, while there is for every other edge only one $\nni$ move that results in a caterpillar tree.
Corresponding to the $\nni$ moves on the lowest internal edge are exchanges of either the first or the second element with the third elements in the sequence representation.

For computing shortest caterpillar paths we modify Bubble Sort as described in the following.
We call this algorithm $\csort$ (Algorithm~\ref{alg:csort}).

\todo{Description + Pseudo code $\csort$}
% But a swap of $c_1$ and $c_2$ does not correspond to any move in $\rnni$.
% Therefore, we need to adapt Bubble Sort for the first three elements of the sequence as follows:
% \todo{LC: add the pseudocode and refer to the line where $\csort$ differs from Bubble Sort! (We assume that we are at the point where we do some changes only on the first positions of the sequence)}
% If the rank of $\parent_R(c_3)$ is higher than the ranks of both $\parent_R(c_1)$ and $\parent_R(c_2)$ in $R$, there are no swaps needed for these taxa.
% If just one of the two taxa $c_1$ or $c_2$ is on the other side of $c_3$ in $R$, this taxon swaps position with $c_3$ in the sequence.
% And if both $c_1$ and $c_2$ are on the other side of $c_3$ in $R$, the taxon that has the parent with higher rank in $R$ swaps with $c_3$.
% This makes sure that there is no additional swap of $c_1$ and $c_2$ needed on the subsequent path to $R$.

@@

In this section we analyse shortest paths in $\rnni$.
Therefore, we will use the algorithms introduced in the previous section.
They enable us to prove that there is always a shortest path between two caterpillar trees that only consists of caterpillar trees (Theorem~\ref{thm:caterpillar_convex}) in Section~\ref{section:caterpillar_convex}.
In other words, the set of caterpillar trees is \emph{convex} in $\rnni$.
